Paper

### 2025-03-14 paper/paper.qmd
intro

# to change body font size:
# change this line in typst-template.typ:
# set text(font: "STIX Two Text", size: 24pt)
library(tidyverse)
library(abc)
library(lhs)
library(flextable)
```

```{r}
# print(here::here())
# file.exists(here::here("scripts", "calibrate.R"))
```

@@

Computer simulation approaches such as agent-based modelling (ABM) can be used to simulate the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact. Agent-based modelling has been used to model various infectious diseases, such as the transmission of important nosocomial pathogens such as methicillin-resistant _Staphylococcus aureus_ (MRSA) within a healthcare setting [@jaramilloAgentBasedModel2015], and the spread of SARS-CoV-2 within a community [@kerrCovasimAgentbasedModel2021].

To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by @salgadoAnalysisOutbreakClostridium2009.

@@

```{r fig-single-outbreak-coloured, include=FALSE}
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| message: false
single_outbreak_sim_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  #slice_sample(n=10) %>% 
  dplyr::select(!`siminputrow`) %>% 
  pivot_longer(cols = !c(`random-seed`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim, `random-seed`) %>%
  summarise(rate = mean(rate)) %>%
  ggplot(aes(x = date_sim,
             y = rate,
             color = factor(`random-seed`))) +
  geom_line() +
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
  guides(color="none")
```

```{r fig-single-outbreak, include=FALSE}
single_outbreak_sim_mod <- single_outbreak_sim %>% 
  mutate(prop_infected = `current-colonised` / `current-inpatients`,
         current_not_infected = `current-inpatients` - `current-colonised`)
single_outbreak_sim_mod %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  ggplot(aes(x = `[step]`, y = value, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment")

# single_outbreak_sum_stat <- single_outbreak_sim_mod %>% 
#   group_by(`[step]`) %>%
#   summarise(m = mean(prop_infected))
```

@@

```{r run-deterministic-model, include=FALSE, cache=TRUE}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

single_seeds <- unique(single_outbreak_sim$`random-seed`)
single_outbreak_one_seed <- single_outbreak_sim_mod %>% 
  filter(`random-seed` == single_seeds[3])

@@

# Time-varying beta function (example: sinusoidal variation over time)
beta_function <- function(t, baseline_rate, outbreak_rate) {
  if (t > unique(single_outbreak_sim$`outbreak-start`) &
      t < unique(single_outbreak_sim$`outbreak-end`)) {
    return (outbreak_rate)
  }
  return (baseline_rate)
}

run_to <- max(single_outbreak_sim$`[step]`)

@@

plan(multisession)
ode_sims <- furrr::future_pmap_dfr(rate_grid,
        \(b_base, b_out, c_base, c_out, r) {
          N <- numeric(length(time))
          I <- numeric(length(time))
          N[1] <- initial_state["N"]
          I[1] <- initial_state["I"]

@@

            # Update populations using Euler's method
            N[t + 1] <- N[t] + dN * dt
            I[t + 1] <- I[t] + dI * dt
          }
          data.frame(baseline = mean(I[31:200] / ( N[31:200] + I[31:200] )),
                     outbreak = mean(I[1100:1200] / ( N[1100:1200] + I[1100:1200] )))
        })

@@

# baselinecolonisation <- 1 - unique(single_outbreak_sim$`community-colonisation-rate`)
# outbreakcolonisation <- 1 - unique(single_outbreak_sim$`o-community-colonisation-rate`)
baselinecolonisation <- mean(abc_single_outbreak$unadj.values[, "c_base"])
outbreakcolonisation <- mean(abc_single_outbreak$unadj.values[, "c_out"])
r <- mean(abc_single_outbreak$unadj.values[, "r"])

@@

```{r fig-single-outbreak-deterministic, include=TRUE}
#| echo: false
single_outbreak_one_seed %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  filter(`[step]` > 31) %>%
  group_by(compartment, `[step]`) %>% 
  summarise(m = mean(value)) %>% 
  ggplot(aes(x = `[step]`, y = m, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment") +
  geom_line(data = output_df, aes(x = time, y = N), color = "blue") + 
  geom_line(data = output_df, aes(x = time, y = I), color = "red") +
  xlim(500, 2000)
```

>>>>

# to change body font size:
# change this line in typst-template.typ:
# set text(font: "STIX Two Text", size: 24pt)
library(tidyverse)
library(abc)
library(lhs)
library(flextable)
library(progressr)
```

@@

Computer simulation approaches such as agent-based modelling (ABM) can be used to simulate the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact. Agent-based modelling has been used to model various infectious diseases, such as the transmission of important nosocomial pathogens such as methicillin-resistant _Staphylococcus aureus_ (MRSA) within a healthcare setting [@jaramilloAgentBasedModel2015], and the spread of SARS-CoV-2 within a community population [@kerrCovasimAgentbasedModel2021]. An important advantage of ABM is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019]. This can be particularly useful in the preparedness for, and management of, outbreaks of healthcare-associated infections, where the impact of interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019]. "What if?" testing allows for different scenarios or bundles of interventions can be simulated to determine their impact.

In this study, we describe an agent-based model that simulates the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. The aim of the model is to guide infection control strategies and interventions, particularly during outbreak periods. The parameters are estimated using data from a CDI outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although the model is designed to be adaptable to other related antimicrobial-resistant pathogens such as VRE, ESBL, and CPE.

@@

```{r fig-single-outbreak-coloured, include=FALSE}
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| message: false
mean_sims_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  #slice_sample(n=10) %>% 
  dplyr::select(!`siminputrow`) %>% 
  pivot_longer(cols = !c(`random-seed`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim, `random-seed`) %>%
  summarise(rate = mean(rate)) %>%
  ggplot(aes(x = date_sim,
             y = rate,
             color = factor(`random-seed`))) +
  geom_line() +
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
  guides(color="none")
```

```{r fig-single-outbreak, include=FALSE}
mean_sims_mod <- mean_sims %>% 
  mutate(prop_infected = `current-colonised` / `current-inpatients`,
         current_not_infected = `current-inpatients` - `current-colonised`)
mean_sims_mod %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  ggplot(aes(x = `[step]`, y = value, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment")

# single_outbreak_sum_stat <- mean_sims_mod %>% 
#   group_by(`[step]`) %>%
#   summarise(m = mean(prop_infected))
```

@@

```{r run-deterministic-model, include=FALSE, cache=TRUE}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(mean_sims$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

single_seeds <- unique(mean_sims$`random-seed`)
single_outbreak_one_seed <- mean_sims_mod %>% 
  filter(`random-seed` == single_seeds[3])

@@

# Time-varying beta function (example: sinusoidal variation over time)
beta_function <- function(t, baseline_rate, outbreak_rate) {
  if (t > unique(mean_sims$`outbreak-start`) &
      t < unique(mean_sims$`outbreak-end`)) {
    return (outbreak_rate)
  }
  return (baseline_rate)
}

run_to <- max(mean_sims$`[step]`)

@@

plan(multisession)
ode_sims <-

    furrr::future_pmap_dfr(rate_grid,
          \(b_base, b_out, c_base, c_out, r) {
            N <- numeric(length(time))
            I <- numeric(length(time))
            N[1] <- initial_state["N"]
            I[1] <- initial_state["I"]

@@

              # Update populations using Euler's method
              N[t + 1] <- N[t] + dN * dt
              I[t + 1] <- I[t] + dI * dt
            }
            data.frame(baseline = mean(I[31:200] / ( N[31:200] + I[31:200] )),
                       outbreak = mean(I[1100:1200] / ( N[1100:1200] + I[1100:1200] )))
          }, .progress = T)

@@

# baselinecolonisation <- 1 - unique(mean_sims$`community-colonisation-rate`)
# outbreakcolonisation <- 1 - unique(mean_sims$`o-community-colonisation-rate`)
baselinecolonisation <- mean(abc_single_outbreak$unadj.values[, "c_base"])
outbreakcolonisation <- mean(abc_single_outbreak$unadj.values[, "c_out"])
r <- mean(abc_single_outbreak$unadj.values[, "r"])

@@

```{r fig-single-outbreak-deterministic-one-plot, include=FALSE}
#| echo: false
#| message: false
#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model. The r"
single_outbreak_one_seed %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  filter(`[step]` > 31) %>%
  group_by(compartment, `[step]`) %>% 
  summarise(m = mean(value)) %>% 
  ggplot(aes(x = `[step]`, y = m, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment") +
  geom_line(data = output_df, aes(x = time, y = N), color = "blue", linetype = "dashed") + 
  geom_line(data = output_df, aes(x = time, y = I), color = "red", linetype = "dashed") +
  xlim(500, 2000)
```

```{r fig-single-outbreak-deterministic-faceted, include=TRUE}
#| echo: false
#| message: false
#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model. The parameters for the deterministic model were determined using approximate Bayesian computation, using the output from a single run of the agent-based model as a reference."
output_df_long <- output_df %>% 
  pivot_longer(cols = c(N, I), names_to = "compartment", values_to = "value") %>% 
  mutate(sim_type = "deterministic") %>% 
  mutate(compartment = case_when(
    compartment == "N" ~ "not_infected",
    compartment == "I" ~ "infected"
  ))

single_outbreak_one_seed %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  filter(`[step]` > 31) %>%
  group_by(compartment, `[step]`) %>% 
  summarise(m = mean(value)) %>% 
  mutate(sim_type = "abm") %>% 
  rename(time = `[step]`, value = m) %>%
  mutate(compartment = case_when(
    compartment == "current_not_infected" ~ "not_infected",
    compartment == "current-colonised" ~ "infected"
  )) %>%
  bind_rows(output_df_long) %>% 
  ggplot(aes(x = time, y = value, color = compartment)) +
  geom_line() +
  facet_wrap(~ sim_type, nrow = 2) +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment")
```
